---
title: "Intro to R, plus some modeling"
author: "Inga Holmdahl"
date: "2/1/2019"
output:
  xaringan::moon_reader:
    css: "slides_style.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Goals for today

1. Make sure everybody has `R`, RStudio, and `deSolve` installed and can run code

--

2. Go through discrete time model of headaches

--

3. Go through non-infectious ODE model of headaches

--

4. Go over questions from headaches worksheet

---

class: center, middle, inverse
## Did anybody here have trouble installing R?

---

class: center, middle, inverse
## Can you run this command with no errors?

```{r eval=FALSE}
library(deSolve)
```
--
If not, make sure you ran `install.packages("deSolve")` first.

---

class: center, middle, inverse
## Do you have the headache.zip file from Canvas?

---
# R Tips

- In `RStudio` highlight the code you want to run in the source panel (upper left) and use **CMD+Enter** or **Ctrl+Enter** to run it in the console

--

- Comment (`#`) your code as you build models (or as Caroline builds them)
    - In a week you'll probably forget why you wrote something some way

--

- Drawing the model and writing the equations first helps (a lot)

--

- We will give you templates for your code

    - You should not be doing this from scratch!

--

- Use `?` to see the arguments for functions

--

- Use **ALT + -** to quickly make an assignment (<-)

--

- It'll be ok! We don't expect you to be `R` masters
---
# Discrete model of headaches

- Here is `headache_discrete.R` with comments removed: 
```{r eval=FALSE}
N_t <- 500      
P_t <- 0        
incidence <- 0.05 
recovery <- 0.9    

dat <- NULL
timesteps <- 1:100

for (t in timesteps){
    N_t1 <- N_t - round(incidence*N_t) + round(recovery*P_t) 
    P_t1 <- P_t + round(incidence*N_t) - round(recovery*P_t)
    dat <- rbind(dat, c(N_t1, P_t1)) 
    
    N_t <- N_t1 
    P_t <- P_t1 
}

matplot(dat, xlab="time", ylab="number of people", type="l", lty = 1)
legend("topright", col = 1:2, legend = c("no headache","headache"), lwd=1)


```

- What's happening in this code? Line-by-line.

---
# Discrete model of headaches

- Examine the output stored in the variable ‘data’, and plot the results. Take a look at the data matrix. What does this tell you about the dynamics of our population?

--

- What does it mean mathematically for the population to be at equilibrium with respect to headaches?

--

- What is the equilibrium prevalence of headaches in the population in this case? 
    - prevalence = 27/473 = 0.057

--

- Verify that **prevalence at equilibrium** is equal to the **incidence x duration**

    - incidence = 0.05
    
    - duration = 1/0.9
    
    - 1/0.9*0.05 = 0.055
--
    - *Note: this approximation only works at low prevalence*

---
# Differential Equation model of headaches

- Where are the differential equations for our model?

``` {r eval=FALSE}

library(deSolve)

times <- seq(0, 5000, by = 1)
yinit <- c(no_headache = 0.95, headache = 0.05)
parameters <- c(incidence = 0.02, recovery = 0.05)

headache_model <- function(times, yinit, parameters) {
    with(as.list(c(yinit, parameters)), {
      
      dno_headache <- recovery*headache - incidence*no_headache 
      dheadache <- incidence*no_headache - recovery*headache
      
      comparts <- list(c(dno_headache, dheadache))
      
      return(comparts)
    })
}

result <- as.data.frame(ode(y = yinit, times = times, 
                            func = headache_model, parms = parameters))

matplot(x = result[, "time"], 
        y = result[, c("no_headache", "headache")], type = "l")
```
---
# Differential Equation model of headaches

- Write them out.

``` {r eval=FALSE}

library(deSolve)

times <- seq(0, 5000, by = 1)
yinit <- c(no_headache = 0.95, headache = 0.05)
parameters <- c(incidence = 0.02, recovery = 0.05)

headache_model <- function(times, yinit, parameters) {
    with(as.list(c(yinit, parameters)), {
      
     {{dno_headache <- recovery*headache - incidence*no_headache}}
     {{dheadache <- incidence*no_headache - recovery*headache}}
      
      comparts <- list(c(dno_headache, dheadache))
      
      return(comparts)
    })
}

result <- as.data.frame(ode(y = yinit, times = times, 
                            func = headache_model, parms = parameters))

matplot(x = result[, "time"], 
        y = result[, c("no_headache", "headache")], type = "l")
```
---

# Differential Equation model of headaches

- How would the model change if we add births and deaths?

``` {r eval=FALSE}

library(deSolve)

times <- seq(0, 5000, by = 1)
yinit <- c(no_headache = 0.95, headache = 0.05)
parameters <- c(incidence = 0.02, recovery = 0.05)

headache_model <- function(times, yinit, parameters) {
    with(as.list(c(yinit, parameters)), {
      dno_headache <- recovery*headache - incidence*no_headache
      dheadache <- incidence*no_headache - recovery*headache
      comparts <- list(c(dno_headache, dheadache))
      return(comparts)
    })
}

result <- as.data.frame(ode(y = yinit, times = times, 
                            func = headache_model, parms = parameters))

matplot(x = result[, "time"], 
        y = result[, c("no_headache", "headache")], type = "l")
```

---

# Differential Equation model of headaches

- How would the model change if we add births and deaths? 
``` {r eval=FALSE}

library(deSolve)

times <- seq(0, 5000, by = 1)
yinit <- c(no_headache = 0.95, headache = 0.05)
{{parameters <- c(incidence = 0.02, recovery = 0.05, b = 0.01, d = 0.01)}}

headache_model <- function(times, yinit, parameters) {
    with(as.list(c(yinit, parameters)), {
      {{dno_headache <- recovery*headache - incidence*no_headache + b - 
       d*dno_headache}}
     {{dheadache <- incidence*no_headache - recovery*headache - d*dheadache}}
      comparts <- list(c(dno_headache, dheadache))
      return(comparts)
    })
}

result <- as.data.frame(ode(y = yinit, times = times, 
                            func = headache_model, parms = parameters))

matplot(x = result[, "time"], 
        y = result[, c("no_headache", "headache")], type = "l")
```
- *Note: we just use `b` here because the total population size is `1`*

---

# Differential Equation model of headaches

- Verify that **prevalence at equilibrium** is equal to the **incidence x duration**
    - incidence = 0.02
    - duration = 1/0.05
    - incidence x duration = 0.4
    
    - prevalence (`result$headache[100]`) = 0.285

- This doesn't really work because of the high prevalence under these parameters

---
class: center, middle, inverse
# Any other questions?